Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-04-27 manuscript.tex
latest

The ideal environment is characterized by the near-absence of stress and disturbance: $d\ll 1$, whereas $b$ is potentially much larger than $1$.

@@

The alternative view is that stressed environments are highly competitive. In spite of inhibited growth and reproduction, population density $N/T$ is actually high since stressful environments also have lower environmental carrying capacity \citep{taylor_1990}. For example, if consumable resources are scarce, we expect intense resource competition (for empirical support, see \citealt{davis_1998}). In our model, territorial contests will then be important.

The mapping of different environments to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}. Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each environment (fourth row; details in Appendix). These can be used to infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fifth row) using a standard invasion analysis.

According to Haldane's formula, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}, with a proportionality factor typically of order one. The fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$). The direction of evolutionary change is determined by the mutational trait changes which are available and also confer an appreciable fitness benefit \citep{yampolsky_2001}, where availability is subject to constraints imposed by the environment.

For example, in Grime's interpretation of stressful environments, $L$ is low, so competition is not important, and only mutants with greater $b$ or lower $d$ will have an appreciably greater $\Delta n_i$. Mutations in $c$ are effectively neutral, and will rarely fix. However, $b$ is constrained to be small. Thus, while some rare mutations may produce small improvements in $b$, it is much more likely that mutations lower $d$, making this the expected direction of evolutionary change.

\begin{figure}
\centering
\begin{tabular}{l*{4}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (HD) \\ \hline
  Constraints & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  Other parameters & $b\gg d$ & $b\gg d$ & $b\approx d$ & $b>d$ \\
  Density $N/T$  & High & Low & Low & High \\
  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c, \downarrow d$
\end{tabular}
\caption{\label{fig:table} The realization of Grime's three environmental extremes in our model, as well as the high-density variant of the stressful environment. Shown are the mapping of each environment to our parameters, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$. *Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq. \eqref{eq:delttot}.}
\end{figure}

@@

and so the propagule density is high $L\approx b_i/d_i\gg 1$, and every unoccupied territory will be heavily contested. The population density is also high $N/T\approx 1$ (since $L=b_iN/(N-T)=b_i/(1-T/N)$).

>>>>

Ideal environments are characterized by the near-absence of stress and disturbance: $d\ll 1$, whereas $b$ could potentially be much larger than $1$.

@@

The alternative view is that stressed environments are highly competitive. In spite of inhibited growth and reproduction, density is actually high since stressful environments also have lower environmental carrying capacity \citep{taylor_1990}. For example, if consumable resources are scarce, we expect intense resource competition (for empirical support, see \citealt{davis_1998}). In our model, territorial contests will then be important.

The mapping of different environments to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}. These can be used to infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fourth row; Appendix C). According to Haldane's formula, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}, with a proportionality factor typically of order one. The fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$). The direction of evolutionary change is determined by the mutational trait changes which are available and also confer an appreciable fitness benefit \citep{yampolsky_2001}, where availability is subject to constraints imposed by the environment. We evaluate the  invasion fitness of mutants under Eq. \eqref{eq:master} for each environmental regime in Appendix C.

\begin{figure}
\centering
\begin{tabular}{l*{4}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (HD) \\ \hline
  Constraints & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  Other parameters & $b\gg d$ & $b\gg d$ & $b\approx d$ & $b>d$ \\
  Density $L$  & $\gg 1$ & $\ll 1$ & $\ll 1$ & $\gg 1$ \\
  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c, \downarrow d$
\end{tabular}
\caption{\label{fig:table} The realization of Grime's three environmental extremes in our model, as well as the high-density variant of the stressful environment. Shown are the mapping of each environment to our parameters, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$. *Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq. \eqref{eq:delttot}.}
\end{figure}

@@

We consider the invasion of mutants into a population with a single type $i$ in equilibrium. The fitness of low frequency invader types $j$ is
\begin{equation}
\Delta n_j/n_j = b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j \label{eq:invad}
\end{equation}
where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$.

In ideal environments, $d$ is much smaller than $1$ whereas $b$ is not. Thus, $1 \ll b_i/d_i = L/(1-e^{-L}) \approx L$, and Eq. \eqref{eq:invad} becomes 
\begin{equation}
\Delta n_j/n_j \approx \frac{b_j c_j}{c_j + L-1}-d_j, \label{eq:invad}
\end{equation}
where without loss of generality we set the reference competitiveness at $c_i=1$.
